---
title: "2019-AUSTINPEAY-1997B Gila Monster Spatial Ecology and Habitat Use"
author: "C.M. Gienger"
date: "April 17, 2020"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_notebook:
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
subtitle: D06 Quarterly Data Deliverable
---

```{r Setup, message=FALSE, include=FALSE}
library(adehabitatHR) #for home range calculations
library(data.table) #manipulate S3 and S4 data tables
library(ggfortify) #to allow ggplot2 to read spatial data
library(grid) #to add annotations to the output
library(knitr) #knitting docs
library(kableExtra) #kable formatted tables
library(leaflet) # quick interactive locator map
library(lubridate) #manipulate date/time
library(rJava) # required for OpenStreetMap
library(OpenStreetMap) #for obtaining raster images
library(patchwork) #making multipanel plots
library(pbapply) #needed for progress bar
library(plotly) #for interactive xy plot
library(rgdal) #for converting spatial data
library(sp) #for converting spatial data
library(tidyverse) #contains dplyr, ggplot2, broom, magrittR, etc.

knitr::opts_chunk$set(fig.path="Figs/", fig.width=6, fig.height=4, echo=FALSE) 
#sends figs in each chunck to folder *after knitting*
```

## Map of Study Site
Calico Basin, Clark County, Nevada (Interactive capability only functional in .html version of document)
```{r Map of Study Site}
leaflet() %>%
  setView(-115.418676, 36.151788, zoom = 13.45) %>% #lat-long of the place of interest
  addProviderTiles(providers$Stamen.Terrain) %>%
  addScaleBar() %>%
  addMiniMap() %>%
  addMarkers(
    lng = -115.418676, lat = 36.151788,
    label = "Calico Basin",
    labelOptions = labelOptions(noHide = T, direction = "right", textOnly = TRUE, textsize = "20px"))
```

```{r Import Data}
data <- read.csv("hesu_bd_20191010.csv")
```

```{r Change Column Names}
data <- data %>% rename(LIZARDNUMBER = HESU., 
                        EASTING = UTM_Easting,
                        NORTHING = UTM_Northing,
                        DATE = Date)
```

```{r Confirm Date Format, results='hide'}
data$DATE <- mdy_hm(data$DATE) #use lubridate to specify *incoming* date format (mdy_hm);
#the current format is "mdy", need to convert column to "ymd" format (IMPOSSIBLE!!!)
str(data$DATE) #look at DATE format to verify that it is converted to "ymd"
```

```{r Summarize Observations at Each Site, results='hide'}
data %>%
  group_by(Site) %>%
  summarize(n())
```

```{r Filter by Study Site}
data <- data %>% filter(Site == "Calico")
```

```{r Add count for number of obs for each lizard to each row of dataframe. Then use the count as a filter}
data <- data %>%
  add_count(LIZARDNUMBER) %>% #adds count to the dataframe
  filter(n > 15)
```

```{r Check the Number of Observations for Each Lizard, results="hide"}
data %>%
  group_by(LIZARDNUMBER) %>%
  count() %>%
  filter(n >= 15) %>%
  arrange(n)
#note that this is just a table to check on the filtering, not a change to the dataframe
```
## Tracking Duration
```{r Tracking Duration, fig.width=8}
#make a chart showing the tracking duration for each individual
ggplot(data, aes(DATE, LIZARDNUMBER)) +
  geom_count (color = "blue", alpha=0.5) + #set symbol size proportional to # of overlapping obs (same day)
  #scale_x_datetime(date_breaks = "12 month", date_labels="%Y")+
  theme(plot.subtitle = element_text(vjust = 1),
        plot.caption = element_text(vjust = 1), 
        axis.text = element_text(size = 12), 
        legend.key = element_rect(fill = NA)) +
  labs(title = "Tracking Duration and Intensity") +
  labs(size = "Number of Obs.")
```

```{r Summarize Temporal Dist. of Observations, even those without UTMs}
data %>%
  group_by(LIZARDNUMBER) %>%
  summarize(n(),
            firstobs = min(DATE),
            lastobs = max(DATE),
            duration = lastobs - firstobs) %>%
  arrange(duration) %>%
  kable(digits =0, #dec. places on duration column
        caption = 'Table 1. Number of telemetry observations (2017-2019).',
        col.names = c("Lizard","Num. of Obs.","First Obs.","Last Obs.","Duration"), 
        align = "c") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

```{r Drop rows with NA data in UTM fields}
data <- data %>% drop_na(EASTING)
```
## Lizard Observations at Calico Basin 
```{r Interactive Plot}
p <- ggplot() + geom_point(data=data, aes(EASTING,NORTHING, color=LIZARDNUMBER)) + labs(x="Easting", y="Northing")
ggplotly(p)
```
This provides the opportunity to look at all data points, or subset to individuals by double-clicking on the ID in the legend. Double-click again to return to all records. (Interactive capability only functional in .html version of document)

```{r Get Raster Imagery, results="hide"}
#CALICO BASIN DATA COLLECTED in NAD83 Zone 11 (WGS84 might be close enough for now)
utm_points <- cbind(data$EASTING, data$NORTHING)
utm_locations <- SpatialPoints(utm_points, proj4string=CRS("+proj=utm +zone=11 +datum=WGS84"))
latlon_locations <- as.data.frame(spTransform(utm_locations, CRS("+proj=longlat +datum=WGS84")))
colnames(latlon_locations) <- c("x","y")
raster <- openmap(c(max(latlon_locations$y)+0.002, min(latlon_locations$x)-0.002), 
                  c(min(latlon_locations$y)-0.002, max(latlon_locations$x)+0.002), type = "bing")
raster_utm <- openproj(raster, projection = "+proj=utm +zone=11 +ellps=WGS84 +units=m +no_defs")
```

```{r Create Default Look, results="hide"}
map_theme <- list(theme(panel.grid.minor = element_blank(),
                       panel.background = element_blank(),
                       plot.background = element_blank(),
                       plot.title = element_blank(),
                       axis.text.x = element_text(vjust = 0.5),
                       axis.title = element_text(face="bold")))
```

```{r Map with Raster Image, echo=FALSE}
autoplot(raster_utm, expand = TRUE) + theme_bw() + theme(legend.position="none") +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
  geom_point(data=data, aes(EASTING,NORTHING, color=LIZARDNUMBER), size = 3, alpha = 0.8) +
  theme(axis.title = element_text(face="bold")) + labs(x="Easting", y="Northing")
```

## Individual Home Ranges and Movements
```{r Faceted Maps, echo=FALSE, fig.width=6, fig.height=8}
autoplot(raster_utm, expand = TRUE) + theme_bw() + theme(legend.position="none") +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
  geom_point(data=data, aes(EASTING,NORTHING, color=LIZARDNUMBER), size = 3, alpha = 0.8) +
  theme(axis.title = element_text(face="bold")) + labs(x="Easting", y="Northing") + facet_wrap(~LIZARDNUMBER)
```
```{r Create Spatial Data, include=FALSE, results="hide"}
x <- as.data.frame(data$EASTING)
y <- as.data.frame(data$NORTHING)
xy <- c(x,y)
data.proj <- SpatialPointsDataFrame(xy,data, proj4string = CRS("+proj=utm +zone=11 +ellps=WGS84 +units=m +no_defs"))
```

```{r message=FALSE, warning=FALSE, results="hide"}
#Without creating additional XY object
mcp(data.proj, percent=100,unout="ha")
```

```{r, results="hide"}
#With additional XY object
xy <- SpatialPoints(data.proj@coords)
mcp.out <- mcp(xy, percent=100, unout="ha")
```

```{r message=FALSE, warning=FALSE, results="hide"}
#Without creating additional XY object
kde<-kernelUD(data.proj, h="href", kern="bivnorm", grid=1000)
ver <- getverticeshr(kde, 95)
```

```{r, results="hide"}
#With additional XY object
kde<-kernelUD(xy, h="href", kern="bivnorm", grid=1000)
ver <- getverticeshr(kde, 95)
```

```{r MCP Points and Polygon, results="hide"}
#mcp plot for all points
mcp.points <- cbind((data.frame(xy)),data$LIZARDNUMBER)
colnames(mcp.points) <- c("x","y", "lizardnumber")
mcp.poly <- fortify(mcp.out, region = "id")
```

```{r KDE Points and Polygon, results="hide"}
#kde plot for all points
kde.points <- cbind((data.frame(data.proj@coords)),data$LIZARDNUMBER)
colnames(kde.points) <- c("x","y","lizardnumber")
kde.poly <- fortify(ver, region = "id")
```

```{r Raster Map with KDE, echo=FALSE, include=FALSE}
autoplot(raster_utm, expand = TRUE) + theme_bw() + theme(legend.position="none") +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
  geom_polygon(data=kde.poly, x=kde.poly$long, y=kde.poly$lat, alpha = 0.35) +
  geom_point(data=kde.points, aes(x=x, y=y, color = lizardnumber)) +
  theme(axis.title = element_text(face="bold")) + labs(x="Easting", y="Northing")
#To create a similar plot for **mcp** simply alter the script; geom_poly() and geom_point() to #values for **mcp.poly** and **mcp.points** objects
```

```{r Generate Individual Files, message=FALSE, warning=FALSE, results="hide"}
#To generate new files we use the following function:
lapply(split(data, data$LIZARDNUMBER), 
       function(x)write.csv(x, file = paste(x$LIZARDNUMBER[1],".csv"), row.names = FALSE))
```
```{r Generate List of Files, results="hide"}
files <- list.files(path = ".", pattern = "[hesuca]+[0-9]", full.names = TRUE)
```

```{r MCP Loop Function, results="hide"}
#looping function for mcp
mcp_analysis <- function(filename){
  data <- read.csv(file = filename)
  x <- as.data.frame(data$EASTING)
  y <- as.data.frame(data$NORTHING)
  xy <- c(x,y)
  data.proj <- SpatialPointsDataFrame(xy,data, proj4string = CRS("+proj=utm +zone=11 +ellps=WGS84 +units=m +no_defs"))
  xy <- SpatialPoints(data.proj@coords)
  mcp.out <- mcp(xy, percent=100, unout="ha")
  area <- as.data.frame(round(mcp.out@data$area,4))
  .rowNamesDF(area, make.names=TRUE) <- data$LIZARDNUMBER
  write.table(area,file="MCP_Hectares.csv", 
              append=TRUE,sep=",", col.names=FALSE, row.names=TRUE)
  mcp.points <- cbind((data.frame(xy)),data$LIZARDNUMBER)
  colnames(mcp.points) <- c("x","y", "lizardnumber")
  mcp.poly <- fortify(mcp.out, region = "id")
  units <- grid.text(paste(round(mcp.out@data$area,2)," ha"), x=0.9,  y=0.95,
                     gp=gpar(fontface=4, cex=0.9), draw = FALSE)
  mcp.plot <- ggplot() +
    geom_polygon(data=mcp.poly, aes(x=mcp.poly$long, y=mcp.poly$lat), alpha=0.5) +
    geom_point(data=mcp.points, aes(x=x, y=y)) + theme_bw() + 
    labs(x="Easting (m)", y="Northing (m)", title=mcp.points$lizardnumber) +
    theme(legend.position="none", plot.title = element_text(face = "bold", hjust = 0.5)) + 
    annotation_custom(units)
  mcp.plot
}
```

```{r KDE Loop Function, results="hide"}
#looping function for kde
kde_analysis <- function(filename){
  data <- read.csv(file = filename)
  x <- as.data.frame(data$EASTING)
  y <- as.data.frame(data$NORTHING)
  xy <- c(x,y)
  data.proj <- SpatialPointsDataFrame(xy,data, proj4string = CRS("+proj=utm +zone=11 +ellps=WGS84 +units=m +no_defs"))
  xy <- SpatialPoints(data.proj@coords)
  kde<-kernelUD(xy, h="href", kern="bivnorm", grid=1000)
  ver <- getverticeshr(kde, 95)
  area <- as.data.frame(round(ver$area,4))
  .rowNamesDF(area, make.names=TRUE) <- data$LIZARDNUMBER
  write.table(area,file="KDE_Hectares.csv", 
              append=TRUE,sep=",", col.names=FALSE, row.names=TRUE)
  kde.points <- cbind((data.frame(data.proj@coords)),data$LIZARDNUMBER)
  colnames(kde.points) <- c("x","y","lizardnumber")
  kde.poly <- fortify(ver, region = "id")
  units <- grid.text(paste(round(ver$area,2)," ha"), x=0.9,  y=0.95,
                     gp=gpar(fontface=4, cex=0.9), draw = FALSE)
  kde.plot <- ggplot() +
    geom_polygon(data=kde.poly, aes(x=kde.poly$long, y=kde.poly$lat), alpha = 0.5) +
    geom_path(data=kde.points, aes(x=x,y=y), linetype = "dashed", colour = "red",
              arrow = arrow(length=unit(0.30,"cm"), angle = 20, ends="last", type = "closed")) +
    geom_point(data=kde.points, aes(x=x, y=y)) + theme_bw() + 
    labs(x="Easting (m)", y="Northing (m)", title=kde.points$lizardnumber) +
    theme(legend.position="none", plot.title = element_text(face = "bold", hjust = 0.5)) + 
    annotation_custom(units)
  kde.plot
}
```

```{r Run KDE with Raster Function on all individuals, results="hide"}
kde_raster <- function(filename){
  data <- read.csv(file = filename)
  x <- as.data.frame(data$EASTING)
  y <- as.data.frame(data$NORTHING)
  xy <- c(x,y)
  data.proj <- SpatialPointsDataFrame(xy,data, proj4string = CRS("+proj=utm +zone=11 +ellps=WGS84 +units=m +no_defs"))
  xy <- SpatialPoints(data.proj@coords)
  kde<-kernelUD(xy, h="href", kern="bivnorm", grid=1000)
  ver <- getverticeshr(kde, 95)
  kde.points <- cbind((data.frame(data.proj@coords)),data$LIZARDNUMBER)
  colnames(kde.points) <- c("x","y","lizardnumber")
  kde.poly <- fortify(ver, region = "id")
  units <- grid.text(paste(round(ver$area,2)," ha"), x=0.9,  y=0.95,
                     gp=gpar(fontface=4, col="white", cex=0.9), draw = FALSE)
  kde.plot <- autoplot(raster_utm, expand = TRUE) + theme_bw() + theme(legend.position="none") +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    geom_polygon(data=kde.poly, aes(x=kde.poly$long, y=kde.poly$lat), alpha = 0.8) +
    geom_point(data=kde.points, aes(x=x, y=y)) +
    labs(x="Easting (m)", y="Northing (m)", title=kde.points$lizardnumber) +
    theme(legend.position="none", plot.title = element_text(face = "bold", hjust = 0.5)) + 
    annotation_custom(units)
  kde.plot
}
```

```{r Examples Individual Home Range Plots, fig.width=8, fig.height=6}
p1 <- mcp_analysis("hesuca3 .csv")
p2 <- kde_analysis("hesuca3 .csv")
p3 <- mcp_analysis("hesuca15 .csv")
p4 <- kde_analysis("hesuca15 .csv")
(p1 + p2) /
  (p3 + p4)
```
```{r Trajectory Loop Function, results="hide"}
#trajectory analysis
traj_analysis <- function(filename){
  relocs_data <- read.csv(file = filename)
  relocs <- as.ltraj(cbind(relocs_data$EASTING, relocs_data$NORTHING),id=relocs_data$LIZARDNUMBER, typeII = FALSE, date=NULL)
  relocs.df <- ld(relocs)
  relocs_dist <- as.data.frame(sum(sapply(relocs.df$dist, sum, na.rm=TRUE)))
  colnames(relocs_dist) <- "Total Distance"
  name <- relocs.df$id[1]
  row.names(relocs_dist) <- name
  relocs_units <- grid.text(paste(round(relocs_dist,2),"m"), x=0.9, y=0.9, 
                            gp=gpar(fontface=3, col="black", cex=0.9), draw = FALSE)
  reloc.plot <- ggplot() + theme_classic() + geom_path(data=relocs.df, aes(x=x,y=y), linetype = "dashed", colour = "red",
    arrow = arrow(length=unit(.5,"cm"), angle = 20, ends="last", type = "closed")) +
    geom_point(data=relocs.df, aes(x=x, y=y)) + geom_point(data=relocs.df, aes(x=x[1], 
    y=y[1]), size = 3, color = "darkgreen", pch=0) +
    labs(x="Easting (m)", y="Northing (m)", title=relocs.df$id[1]) +
    theme(legend.position="none", plot.title = element_text(face = "bold", hjust = 0.5)) +
    annotation_custom(relocs_units)
  reloc.plot
}
```

```{r Distance over time Loop Function, results="hide"}
#distance over time analysis
dist_analysis <- function(filename){
  relocs_data <- read.csv(file = filename)
  relocs <- as.ltraj(cbind(relocs_data$EASTING, relocs_data$NORTHING),id=relocs_data$LIZARDNUMBER, typeII = FALSE, date=NULL)
  relocs.df <- ld(relocs)
  relocs_dist <- as.data.frame(sum(sapply(relocs.df$dist, sum, na.rm=TRUE)))
  colnames(relocs_dist) <- "Total Distance"
  name <- relocs.df$id[1]
  row.names(relocs_dist) <- name
  write.table(relocs_dist,file="reloc_dist.csv", 
              append=TRUE,sep=",", col.names=FALSE, row.names=TRUE)
  dist.plot <- ggplot() + geom_point(data = relocs.df, aes(x=date, y=dist), na.rm = TRUE) + theme_classic() +
    geom_path(data = relocs.df, aes(x=date, y=dist), stat = "identity", na.rm = TRUE) +
    labs(x="Observation Number", y="Distance Traveled (m)", title=relocs.df$id[1]) + 
    theme(legend.position="none", plot.title = element_text(face = "bold", hjust = 0.5))
  dist.plot
}
```
## Trajectories of Movement and Distances Traveled
```{r Examples Individual Relocation Plots, fig.width=8, fig.height=6}
p5 <- traj_analysis("hesuca3 .csv")
p6 <- traj_analysis("hesuca15 .csv")
p7 <- dist_analysis("hesuca3 .csv")
p8 <- dist_analysis("hesuca15 .csv")

(p5 + p6)/
  (p7 + p8)
```







































